---
title: "Cost Analysis for Shane's Candies"
author: "Allocate Analytics"
date: "August 19, 2018"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Read in functions ###

s.gsheet.imp <- function(googlesheet.name){
  library(googlesheets)
  gs_auth()
  source.gsheet <- gs_title(googlesheet.name)
  imported.data <- gs_read(source.gsheet)#, range = "A1:AQ100")
  imported.data <- as.data.frame(imported.data)
  print(imported.data)
  imported.data
}

unit.converter <- function(df, quantity.col.num, units.col.num){
  conversions <- data.frame("eng.meas" = c("unit(s)", "ounce(s) fl","cup(s) fl", "quart(s)", "gallon(s)", "tablespoon(s) fl", "cup(s) dry", "tablespoon(s) dry", "ounce(s) dry", "lb(s)"),
                            "to.metric" = c(1, 29.57, 240, 15, 946.35, 3840, 190, 12, 30, 453.59),
                            "metric.unit" = c("units", "ml","ml", "ml", "ml", "ml", "g", "g", "g", "g"))
  library(dplyr)
  
  #print(conversions)
  
  num.convert <- function(quantity.col, eng.units.col){
    this.conversion <- conversions %>%
      filter(eng.meas == eng.units.col)
    #print(this.conversion)
    quantity.col <- quantity.col*this.conversion$to.metric
  }
  
  num.col <- as.numeric(mapply(num.convert, df[,quantity.col.num], df[,units.col.num]))
  #print(num.col)
  
  unit.convert <- function(eng.units.col){
    this.conversion <- conversions[conversions$eng.meas == eng.units.col,]
    eng.units.col <- as.character(this.conversion$metric.unit)
    # use as.character() on the result after sapply
    eng.units.col
  }
  
  units.col <- as.character(sapply(df[,units.col.num], unit.convert))
  print(units.col)
  
  df[,quantity.col.num] <- num.col
  df[,units.col.num] <- units.col

  #print(df)
  df
}

costs.per.unit <- function(unit.converted.cost.list.df, price.col.num, quantity.col.num){
  cost.df <- unit.converted.cost.list.df
  unit.divide <- function(a,b){a <- (a/b)}
  price.per.unit.col <- mapply(unit.divide, cost.df[,price.col.num], cost.df[,quantity.col.num])
  #quantity.one.col <- 1
  
  cost.df[,price.col.num] <- price.per.unit.col
  cost.df[,quantity.col.num] <- 1
  cost.df
}

reshape.costs.ing <- function(imported.ingredient.info.df, raw.unit.costs.df, rows.of.products){
  conv.df <- imported.ingredient.info.df
  
  unit.converter <- function(df, quantity.col.num, units.col.num){
    conversions <- data.frame("eng.meas" = c("unit(s)", "ounce(s) fl","cup(s) fl", "quart(s)", "gallon(s)", "tablespoon(s) fl", "cup(s) dry", "tablespoon(s) dry", "ounce(s) dry", "lb(s)"),
                              "to.metric" = c(1, 29.57, 240, 15, 946.35, 3840, 190, 12, 30, 453.59),
                              "metric.unit" = c("units", "ml","ml", "ml", "ml", "ml", "g", "g", "g", "g"))
    library(dplyr)
    
    print(conversions)
    
    num.convert <- function(quantity.col, eng.units.col){
      this.conversion <- conversions %>%
        filter(eng.meas == eng.units.col)
      #print(this.conversion)
      quantity.col <- quantity.col*this.conversion$to.metric
    }
    
    num.col <- as.numeric(mapply(num.convert, df[,quantity.col.num], df[,units.col.num]))
    print(num.col)
    
    unit.convert <- function(eng.units.col){
      this.conversion <- conversions[conversions$eng.meas == eng.units.col,]
      eng.units.col <- as.character(this.conversion$metric.unit)
      # use as.character() on the result after sapply
      eng.units.col
    }
    
    units.col <- as.character(sapply(df[,units.col.num], unit.convert))
    print(units.col)
    
    df[,quantity.col.num] <- num.col
    df[,units.col.num] <- units.col
    
    print(df)
    df
  }
  
  colnames(conv.df)[3:32] <- rep(c("Ing.or.Phase", "Quantity", "Unit"), 10)
  all.long.df <- data.frame()
  for(i in rows.of.products){
    print(i)
    if(!is.na(conv.df[i,1])){
  ingreds <- rbind(conv.df[i,3:5], conv.df[i,6:8], conv.df[i,9:11], conv.df[i, 12:14], 
                   conv.df[i, 15:17], conv.df[i, 18:20], conv.df[i, 21:23], conv.df[i, 24:26]
                   , conv.df[i, 27:29], conv.df[i, 30:32])
  ingreds <- ingreds[!is.na(ingreds$Unit),]
  long.df <- cbind("Product" = conv.df[i,1], "Batch.Size" = conv.df[i,2], "Cost.Type" = "Ingredient", ingreds)
  all.long.df <- rbind(all.long.df, long.df)
    } #ends the NA check
  }
  
  all.long.df <- unit.converter(all.long.df,5,6)
  
  all.long.merged <- merge(all.long.df, raw.unit.costs.df[,1:2], by.x = "Ing.or.Phase", by.y = "Ingredient Name", 
                           all.x = TRUE, all.y = FALSE)
  all.long.merged$Cost <- all.long.merged$`Price Paid`*all.long.merged$Quantity
  all.long.merged$Batch.Size <- as.numeric(all.long.merged$Batch.Size)
  all.long.merged$Cost.Div.Batch <- all.long.merged$Cost/all.long.merged$Batch.Size
  all.long.merged <- all.long.merged %>%
    select(Product, Batch.Size, Cost.Type, Ing.or.Phase, Quantity, Unit, Cost, Cost.Div.Batch) %>%
    arrange(Product)
  print(all.long.merged)
  all.long.merged

}

reshape.costs.labor <- function(imported.ingredient.info.df, rows.of.products){
  library(dplyr)

  conv.df <- imported.ingredient.info.df
  temp.col.names <- rep(c("Length", "People"), 5)
  colnames(conv.df)[34:43] <- temp.col.names
  
  #colnames(conv.df)[3:32] <- rep(c("Ing.or.Phase", "Quantity", "Units"), 10)
  all.long.df <- data.frame()
  for(i in rows.of.products){
    print(i)
    if(!is.na(conv.df[i,1])){
      labors <- rbind(conv.df[i,34:35], conv.df[i,36:37], conv.df[i,38:39], conv.df[i, 40:41], 
                       conv.df[i, 42:43])
      labor.hours.df <- cbind(labors, labor.hours = labors[,1]*labors[,2])

      labor.hours.df <- labor.hours.df[!is.na(labor.hours.df$labor.hours),]
      print(labor.hours.df)
      long.df <- cbind("Product" = conv.df[i,1], "Batch.Size" = conv.df[i,2], 
                       "Cost.Type" = "Labor", "Ing.or.Phase" = paste("Phase", seq_len(nrow(labor.hours.df))),
                       "Quantity" = labor.hours.df$labor.hours, "Unit" = "labor hours")
      long.df <- as.data.frame(long.df)

      pay.rate <- as.numeric(conv.df[i,33])
      long.df$Quantity <- as.numeric(long.df$Quantity)
      long.df$Cost <- long.df$Quantity*pay.rate
      long.df$Batch.Size <- as.numeric(as.vector(long.df$Batch.Size))
      #print(head(long.df$Batch.Size, 20))
      long.df$Cost.Div.Batch <- long.df$Cost/long.df$Batch.Size
      print(long.df)      
      all.long.df <- rbind(all.long.df, long.df)
    } #ends the NA check
  }
  
  print(all.long.df)
  all.long.df
}

profit.calc.append <- function(combined.df, imported.retail.price.df){
  print("need df with both ingredients and phases which includes Cost.Div.Batch column")
  
  cost.agg <- aggregate(Cost.Div.Batch ~ Product + Batch.Size, combined.df, sum)
  print(cost.agg)
  
  price.df <- imported.retail.price.df
  print(str(price.df))

  cost.price <- merge(cost.agg, price.df, all.x=TRUE, all.y=FALSE)
  cost.price <- cost.price %>%
    mutate("Profit" = `Retail Price`-Cost.Div.Batch)
  print(cost.price)
  new.rows <- as.data.frame(cbind("Product" = as.character(cost.price$Product), "Batch.Size" = cost.price$Batch.Size, "Cost.Type" = "Profit", "Ing.or.Phase" = NA, "Quantity" = 1, "Unit" = "units", "Cost" = (cost.price$Profit*cost.price$Batch.Size), "Cost.Div.Batch" = cost.price$Profit))
  print(colnames(new.rows))
  combined.df <- rbind(combined.df, new.rows)

}

component.plot2 <- function(df, product.to.plot, palette.vector){
  
  #df <- arrange(df, Ing.or.Phase)
  df$Ing.or.Phase <- factor(df$Ing.or.Phase, 
                            levels = c("Phase 8", "Phase 7", "Phase 6", "Phase 5", 
                                       "Phase 4", "Phase 3", "Phase 2", "Phase 1", 
                                       "Cocoa", "Sugar", "Milk", "Cashews", "Frosting", 
                                       "Eggs", "Flour", "Caramel", "Vanilla bean", 
                                       "Liquid Sugar", "Lollipop sticks", "Wrapper"))
  library(ggplot2)
  library(scales)
  library(dplyr)
  
  #print(df)
  #print(str(df))
  bar.plot1 <- ggplot(data=filter(df, Product == product.to.plot), 
                      aes(x=Ing.or.Phase, y=Cost.Div.Batch, group = Cost.Type)) + 
    geom_col(aes(fill = Cost.Type), position = "dodge") +
    #geom_label(aes(label = factor(Ing.or.Phase)), position = position_stack(vjust = 0.5))+
    scale_y_continuous(labels = scales::dollar) +
    scale_fill_manual(values = palette.vector) +
    #theme(legend.position = "none") +
    labs(x= "Cost Type", y="Cost in One Unit", title = paste0("Cost Breakdown for ", product.to.plot)) +
    theme_dark() +
    theme(legend.title = element_blank()) +
    coord_flip() +
    NULL
  
  bar.plot1
}

full.stack.plot <- function(df, product.to.plot, palette.vector){
  
  library(ggplot2)
  library(scales)
  library(dplyr)

  df$Cost.Div.Batch <- as.numeric(as.vector(df$Cost.Div.Batch))
  
  df.agg <- aggregate(Cost.Div.Batch ~ Product + Cost.Type, df, sum)
  
  df.agg$Cost.Type <- factor(df.agg$Cost.Type, levels = c("Labor", "Ingredient", "Profit"))
  bar.plot1 <- 0

  bar.plot1 <- ggplot(data=filter(df.agg, Product == product.to.plot), aes(x=Product, y=Cost.Div.Batch)) + 
    #bar.plot1 <- ggplot(data=df.agg, aes(x=Product, y=Cost.Div.Batch)) + 
      geom_col(aes(fill = factor(Cost.Type, levels = c("Labor", "Ingredient", "Profit"))), 
               position = position_stack(reverse = TRUE), width = .4) +
      #geom_label(aes(label = factor(Cost.Type, levels = c("Labor", "Ingredient", "Profit"))), 
      #               position = position_stack(reverse = FALSE, vjust = 0.5))+
      scale_y_continuous(labels = scales::dollar) +
      scale_fill_manual(values = palette.vector) +
      labs(x="Product", y="Costs and Profit in Retail Price", subtitle = "A green bar below zero indicates a loss", title = paste0("Costs and Profit for ", product.to.plot)) +
      theme_dark() +
    theme(legend.title = element_blank()) +
    NULL
  
  bar.plot1
}

# just read in the products for titles
product.ingredients <- s.gsheet.imp("Fake SC Ingredient Data")

product.vec <- product.ingredients[,1]

### Process Data ###

# import store's cost to purchase ingredients:
ingredient.costs <- s.gsheet.imp("Fake SC Ingredient Costs") 

# convert to metric:
ing.costs.metric <- unit.converter(ingredient.costs, quantity.col.num = 3, units.col.num = 4) 

# convert to one metric unit:
ing.costs.metric.one <- costs.per.unit(ing.costs.metric, price.col.num = 2, quantity.col.num = 3) 

#import data on which ingredients are in products:
ingredient.data <- s.gsheet.imp("Fake SC Ingredient Data")

ing.data.long <- reshape.costs.ing(ingredient.data, ing.costs.metric.one, rows.of.products = 1:9)

labor.data.long <- reshape.costs.labor(ingredient.data, rows.of.products = 1:9)

combined.data.long <- rbind(ing.data.long, labor.data.long)

retail.price.of.products <- s.gsheet.imp("Fake SC Pricing Info")

combined.data.long.prof <- profit.calc.append(combined.data.long, retail.price.of.products)


next.product <- product.vec[1] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

These charts show the cost components of the top selling products.  The chocolate color shows the physical ingredients and the vanilla color shows the labor.  The first chart focuses on individual ingredients and phases of production while the second one focuses on what portion of the total retail price is comprised of ingredients, labor, and profit.  In this version fake data was created for 1) ingredients in each product, 2) bulk costs of ingredients, and 3) retail prices of individual products.

```{r product1, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

########
next.product <- product.vec[2] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product2, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[3] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product3, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[4] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product4, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[5] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product5, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[6] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product6, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[7] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product7, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[8] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product8, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

next.product <- product.vec[9] # this avoids having to change the subset number in multiple places 
```

# Cost and Margin for `r next.product`

```{r product9, echo=FALSE, message=FALSE, warning=FALSE}
choc.green <- c("#6F2919", "#E8D5A1", "#98F95D")

component.plot2(combined.data.long, next.product, choc.green)

choc.green.swap <- c("#E8D5A1", "#6F2919", "#98F95D")

full.stack.plot(combined.data.long.prof, next.product, choc.green.swap)

#next.product <- product.vec[2] # this avoids having to change the subset number in multiple places 
```